MassWateR: Improving quality control, analysis, and sharing of water quality data

### 2023-07-28 paper.tex
edits for BW comments done

% !TeX root = RJwrapper.tex
\title{MassWateR: Improving quality control, analysis, and sharing of water quality data}
\author{by Marcus W. Beck, Benjamen Wetherill, Jillian Carr, and Pamela DiBona}

@@

Monitoring data of sufficient quantity and quality are critical to ensure precise and accurate representation of environmental conditions. A significant bottleneck in the use of monitoring data for environmental assessment of surface waters is the ability to clearly and efficiently indicate that the data fulfill applicable QC checks for regulatory applications or inclusion in a consolidated database (Arndt et al. 2022). Common QC checks for \emph{in situ} field measurements or concentrations measured in the laboratory may include 1) comparison of the precision between replicate samples (duplicates), 2) comparison of a sample to a known concentration (spikes or instrument checks), and 3) precision of the measurement from an empty or blank sample (blanks) (Wilde and U.S. Geological Survey 2002). An adequate number of QC samples must also be included in the dataset as a measure of ``completeness''. These checks are often compiled in a single report for review by appropriate regulatory agencies. For example, precision of duplicate samples for a given parameter must not vary more than 5\% and at least 10\% of the data should be dedicated to these checks as a measure of completeness. For local monitoring groups that lack the resources to develop robust and repeatable workflows, QC reports are often prepared manually before submitting the data for review. This process is time-consuming and prone to errors, often limiting the amount of useful information for regulatory assessments or submitted to formal databases.

@@

To our knowledge, there are no existing R packages on CRAN that can be used to facilitate QC of water quality data, nor are any available that facilitate submission to existing databases. However, there are several that can be used to retrieve and analysis data from existing sources (see the CRAN \ctv{Hydrology} Task View). In particular, the \CRANpkg{dataRetrieval} package (De Cicco et al. 2022) has been used widely to retrieve data from the USEPA Water Quality Portal (WQP), which is the counterpart of the WQX system for accessing data submitted using the latter. This package leverages a robust API to query existing water quality data in standardized format provided by the WQP. As such, data retrieval using existing web services is much simpler than data submission, as data formatting requirements do not apply when retrieving data. Developing a robust tool that can facilitate the upload of data to WQX, in addition to streamlining QC processes, would further the value of packages like \CRANpkg{dataRetrieval} by increasing the amount of data that can be accessed through the WQP. The \CRANpkg{MassWateR} package was developed to provide this benefit.

Users can engage with \CRANpkg{MassWateR} to achieve different goals. This design was intentional based on likely differences in needs among the user community. Although increasing data submission and facilitating QC reporting was the primary goal, we also assumed that users may not want to do both. That is, QC reporting is not a requirement to submit to WQX, whereas state institutions require this reporting for regulatory assessment. Users may also simply have a need to understand trends or to summarize their data, while also wanting to extend these analyses beyond \CRANpkg{MassWateR} using additional R packages. Figure \ref{fig:workflow} demonstrates how a user may apply the functions in \CRANpkg{MassWateR} once the required data are imported. The functions allow a user to engage with their data several ways, including 1) screening data for quality control, 2) summarizing quality control results into a single report or separate tables, 3) creating graphics for analysis and reports to stakeholders, and 4) formatting data for upload to WQX.

@@

The required data files for using \CRANpkg{MassWateR} are shown in Table \ref{tab:filreq}, including the files that apply to the workflow steps in Figure \ref{fig:workflow}. The files are each imported into R using specific \texttt{read} functions with relevant checks, explained in the next section. These checks verify multiple requirements outlined in the template files, with informative errors or warnings returned to the console to prompt the user on the required action to remedy a formatting issue. The largest input file required for all parts of the workflow in Figure \ref{fig:workflow} is the results file. This file includes all water quality monitoring data to be used with the package. As such, the formatting requirements for the results file are the most burdensome for potential users and additional functions are available to assist in this effort.

@@

\begin{verbatim}
#> Running checks on results data...
\end{verbatim}

\begin{verbatim}
#>  Checking column names... OK
\end{verbatim}

\begin{verbatim}
#>  Checking all required columns are present... OK
\end{verbatim}

\begin{verbatim}
#>  Checking valid Activity Types... OK
\end{verbatim}

\begin{verbatim}
#>  Checking Activity Start Date formats... OK
\end{verbatim}

\begin{verbatim}
#>  Checking depth data present... OK
\end{verbatim}

\begin{verbatim}
#>  Checking for non-numeric values in Activity Depth/Height Measure... OK
\end{verbatim}

\begin{verbatim}
#>  Checking Activity Depth/Height Unit... OK
\end{verbatim}

\begin{verbatim}
#>  Checking Activity Relative Depth Name formats... OK
\end{verbatim}

\begin{verbatim}
#>  Checking values in Activity Depth/Height Measure > 1 m / 3.3 ft... OK
\end{verbatim}

\begin{verbatim}
#>  Checking Characteristic Name formats... OK
\end{verbatim}

\begin{verbatim}
#>  Checking Result Values... OK
\end{verbatim}

\begin{verbatim}
#>  Checking QC Reference Values... OK
\end{verbatim}

\begin{verbatim}
#>  Checking for missing entries for Result Unit... OK
\end{verbatim}

\begin{verbatim}
#>  Checking if more than one unit per Characteristic Name... OK
\end{verbatim}

\begin{verbatim}
#>  Checking acceptable units for each entry in Characteristic Name... OK
\end{verbatim}

\begin{verbatim}
#> 
#> All checks passed!
\end{verbatim}

@@

\begin{verbatim}
#> Running checks on results data...
\end{verbatim}

\begin{verbatim}
#>  Checking column names... OK
\end{verbatim}

\begin{verbatim}
#>  Checking all required columns are present... OK
\end{verbatim}

\begin{verbatim}
#> Error:   Checking valid Activity Types...
#>  Incorrect Activity Type found: Sample, Field in row(s) 4, 135
\end{verbatim}

The \texttt{readMWRresultsview()} function is also available to assist with troubleshooting formatting issues for the results file. This function exports an Excel file that shows the unique values that are found in each column to allow a user to quickly see potential incorrect entries. The output is similar to running \texttt{apply(resdat,\ 2,\ unique)} in the console, but an external file is created that users may be more comfortable evaluating to troubleshoot formatting problems.

@@

A final note about the data inputs is that many types of water quality measurements can be included for analysis, although the package is currently limited to working with discrete samples as compared to data from continuous monitoring equipment. The \texttt{paramsMWR} dataset included with \CRANpkg{MassWateR} provides a complete table of the parameters that can be used with the package. On data import, this dataset is referenced to ensure that only relevant parameters are included and that appropriate units of measurement are provided. The dataset includes 43 different parameters, each with multiple valid units of measurement. Additionally, only one unit of measurement is allowed per parameter, which was an intentional design so that tedious functions for converting between dozens of units of measurement did not need to be created during package development. It is not an unreasonable expectation for users to provide only one unit of measurement per parameter. Laboratories typically have a standard reporting format based on the same methodology or instrument used to measure concentrations of water quality parameters.

An additional set of functions can be used to check for outliers in the results file. Outliers can be caused by several reasons, including data entry errors, laboratory processing errors, or anomalous environmental events. Identifying and potentially removing or correcting outliers is a critical part of quality control. The functions in \CRANpkg{MassWateR} simply identify potential outliers to provide an opportunity for users to address these values. The decision on how to address outliers remains with the user and no automated tools are provided in the package to correct or remove outliers.

@@

\begin{center}\includegraphics[width=0.95\linewidth]{paper_files/figure-latex/unnamed-chunk-4-1} \end{center}

@@

@@

The QC report is built using several functions that can be used individually as needed. In particular, the \texttt{tabMWRacc()}, \texttt{tabMWRfre()}, and \texttt{tabMWRcom()} create \CRANpkg{flextable} (Gohel and Skintzos 2023) objects that can be viewed in RStudio and are compatible with Microsoft Word output. These functions are useful for understanding how the QC checks are created for the separate components of the QC report. For example, the \texttt{tabMWRacc()} function evaluates accuracy checks for field duplicates, lab duplicates, field blanks, lab blanks, and lab spikes/instrument checks for QC records in the results file based on DQOs in the accuracy file. The function can return a summary of all checks as follows:

@@

\begin{longtable}[c]{|p{1.47in}|p{0.82in}|p{1.08in}|p{0.88in}|p{0.72in}}

@@

\multicolumn{1}{>{\raggedright}m{\dimexpr 1.47in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{Type}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedright}m{\dimexpr 0.82in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{Parameter}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedright}m{\dimexpr 1.08in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{Number\ of\ QC\ Checks}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedright}m{\dimexpr 0.88in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{Number\ of\ Misses}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedright}m{\dimexpr 0.72in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\%\ Acceptance}}} \\

\ascline{1.5pt}{666666}{1-5}\endfirsthead

\multicolumn{1}{>{\raggedright}m{\dimexpr 1.47in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{Type}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedright}m{\dimexpr 0.82in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{Parameter}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedright}m{\dimexpr 1.08in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{Number\ of\ QC\ Checks}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedright}m{\dimexpr 0.88in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{Number\ of\ Misses}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedright}m{\dimexpr 0.72in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\%\ Acceptance}}} \\

@@

\multicolumn{1}{>{\raggedright}m{\dimexpr 1.47in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{Field\ Duplicates}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedright}m{\dimexpr 0.82in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedright}m{\dimexpr 1.08in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedright}m{\dimexpr 0.88in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedright}m{\dimexpr 0.72in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{}}} \\

@@

\multicolumn{1}{>{\raggedright}m{\dimexpr 1.47in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedright}m{\dimexpr 0.82in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{Ammonia}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedright}m{\dimexpr 1.08in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{4}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedright}m{\dimexpr 0.88in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{1}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedright}m{\dimexpr 0.72in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{75\ \%}}} \\

@@

\multicolumn{1}{>{\raggedright}m{\dimexpr 1.47in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedright}m{\dimexpr 0.82in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{DO}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedright}m{\dimexpr 1.08in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{11}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedright}m{\dimexpr 0.88in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{0}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedright}m{\dimexpr 0.72in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{100\ \%}}} \\

@@

\multicolumn{1}{>{\raggedright}m{\dimexpr 1.47in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedright}m{\dimexpr 0.82in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{E.coli}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedright}m{\dimexpr 1.08in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{2}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedright}m{\dimexpr 0.88in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{0}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedright}m{\dimexpr 0.72in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{100\ \%}}} \\

@@

\multicolumn{1}{>{\raggedright}m{\dimexpr 1.47in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedright}m{\dimexpr 0.82in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{Nitrate}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedright}m{\dimexpr 1.08in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{2}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedright}m{\dimexpr 0.88in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{0}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedright}m{\dimexpr 0.72in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{100\ \%}}} \\

@@

\multicolumn{1}{>{\raggedright}m{\dimexpr 1.47in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedright}m{\dimexpr 0.82in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{pH}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedright}m{\dimexpr 1.08in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{11}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedright}m{\dimexpr 0.88in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{0}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedright}m{\dimexpr 0.72in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{100\ \%}}} \\

@@

\begin{longtable}[c]{|p{0.55in}|p{0.60in}|p{0.27in}|p{0.50in}|p{0.69in}|p{0.45in}}

\caption{Field\ Blanks}\label{tab:unnamed-chunk-10}\\

@@

\begin{center}\includegraphics[width=0.95\linewidth]{paper_files/figure-latex/unnamed-chunk-11-1} \end{center}

@@

>>>>

% !TeX root = RJwrapper.tex
\title{MassWateR: Improving quality control, analysis, and sharing of water quality data}
\author{by Marcus W. Beck, Benjamen Wetherill, and Jillian Carr}

@@

Monitoring data of sufficient quantity and quality are critical to ensure precise and accurate representation of environmental conditions. A significant bottleneck in the use of monitoring data for environmental assessment of surface waters is the ability to clearly and efficiently indicate that the data fulfill applicable QC checks for regulatory applications or inclusion in a consolidated database (Arndt et al. 2022). Common QC checks for \emph{in situ} field measurements or concentrations measured in the laboratory may include 1) comparison of the precision between replicate samples (duplicates), 2) accuracy of a sample relative to a known concentration (spikes or instrument checks), and 3) evaluation of the measurement from an empty or blank sample (blanks) (Wilde and U.S. Geological Survey 2002). An adequate number of QC samples must also be included in the dataset as a measure of ``completeness''. These checks are often compiled in a single report for review by appropriate regulatory agencies. For example, precision of duplicate samples for a given parameter must not vary more than 5\% and at least 10\% of the data should be dedicated to these checks as a measure of completeness. For local monitoring groups that lack the resources to develop robust and repeatable workflows, QC reports are often prepared manually before submitting the data for review. This process is time-consuming and prone to errors, often limiting the amount of useful information for regulatory assessments or submitted to formal databases.

@@

To our knowledge, there are no existing R packages on CRAN that can be used to facilitate QC of water quality data, nor are any available that facilitate submission to existing databases. However, there are several that can be used to retrieve and analyze data from existing sources (see the CRAN \ctv{Hydrology} Task View). In particular, the \CRANpkg{dataRetrieval} package (De Cicco et al. 2022) has been used widely to retrieve data from the USEPA Water Quality Portal (WQP), which is the counterpart of the WQX system for accessing data submitted using the latter. This package leverages a robust API to query existing water quality data in standardized format provided by the WQP. As such, data retrieval using existing web services is much simpler than data submission, as data formatting requirements do not apply when retrieving data. Developing a robust tool that can facilitate the upload of data to WQX, in addition to streamlining QC processes, would further the value of packages like \CRANpkg{dataRetrieval} by increasing the amount of data that can be accessed through the WQP. The \CRANpkg{MassWateR} package was developed to provide this benefit.

Users can engage with \CRANpkg{MassWateR} to achieve different goals. This design was intentional based on likely differences in needs among the user community. Although increasing data submission and facilitating QC reporting was the primary goal, we also assumed that users may not want to do both. That is, state institutions require QC reporting for regulatory assessment, whereas data submission to WQX may be a separate process. Users may also simply have a need to understand trends or to summarize their data, while also wanting to extend these analyses beyond \CRANpkg{MassWateR} using additional R packages. Figure \ref{fig:workflow} demonstrates how a user may apply the functions in \CRANpkg{MassWateR} once the required data are imported. The functions allow a user to engage with their data several ways, including 1) screening data for quality control, 2) summarizing quality control results into a single report or separate tables, 3) creating graphics for analysis and reports to stakeholders, and 4) formatting data for upload to WQX.

@@

The required data files for using \CRANpkg{MassWateR} are shown in Table \ref{tab:filreq}, including the files that apply to the workflow steps in Figure \ref{fig:workflow}. The files are each imported into R using specific \texttt{read} functions with relevant checks, explained in the next section. These checks verify multiple requirements outlined in the template files, with informative errors or warnings returned to the console to prompt the user on the required action to remedy a formatting issue. The largest input file required for all parts of the workflow in Figure \ref{fig:workflow} is the results file. This file includes all water quality monitoring data to be used with the package and the recommendation is that the results data include a year of sampling as reporting typically follows an annual cycle. As such, the formatting requirements for the results file are the most burdensome for potential users and additional functions are available to assist in this effort.

@@

\begin{verbatim}
#> Running checks on results data...
#>    Checking column names... OK
#>    Checking all required columns are present... OK
#>    Checking valid Activity Types... OK
#>    Checking Activity Start Date formats... OK
#>    Checking depth data present... OK
#>    Checking for non-numeric values in Activity Depth/Height Measure... OK
#>    Checking Activity Depth/Height Unit... OK
#>    Checking Activity Relative Depth Name formats... OK
#>    Checking values in Activity Depth/Height Measure > 1 m / 3.3 ft... OK
#>    Checking Characteristic Name formats... OK
#>    Checking Result Values... OK
#>    Checking QC Reference Values... OK
#>    Checking for missing entries for Result Unit... OK
#>    Checking if more than one unit per Characteristic Name... OK
#>    Checking acceptable units for each entry in Characteristic Name... OK
#> All checks passed!
\end{verbatim}

@@

\begin{verbatim}
#> Running checks on results data...
#>    Checking column names... OK
#>    Checking all required columns are present... OK
#> Error:  Checking valid Activity Types...
#>    Incorrect Activity Type found: Sample, Field in row(s) 4, 135
\end{verbatim}

The \texttt{readMWRresultsview()} function is also available to assist with troubleshooting formatting issues for the results file. This function exports an Excel file that shows the unique values that are found in each column to allow a user to quickly see potential incorrect entries. The output is similar to running the following:

\begin{verbatim}
apply(readxl::read_excel('path/to/results.xlsx'), 2, unique)
\end{verbatim}

However, this approach returns unique column entries in the R console as a list object. Users may be more comfortable evaluating the external file created by \texttt{readMWRresultsview()} to troubleshoot formatting problems.

@@

A final note about the data inputs is that many types of water quality measurements can be included for analysis, although the package is currently limited to working with discrete samples as compared to data from continuous monitoring equipment. The \texttt{paramsMWR} dataset included with \CRANpkg{MassWateR} provides a complete table of the parameters that can be used with the package. On data import, this dataset is referenced to ensure that only relevant parameters are included and that appropriate units of measurement are provided. The dataset includes 43 different parameters, each with multiple valid units of measurement. Additionally, only one unit of measurement is allowed per parameter, which was an intentional design so that tedious functions for converting between dozens of units of measurement did not need to be created during package development. It is not an unreasonable expectation for users to provide only one unit of measurement per parameter. Laboratories typically have a standard reporting format based on the same methodology or instrument used to measure concentrations of water quality parameters. Both the number of parameters and acceptable units can easily be extended and those herein were chosen simply to limit the development of \CRANpkg{MassWateR} to a manageable scope.

\hypertarget{outliers}{%
\subsection{Outliers}\label{outliers}}

Checking the results file for outliers is an additional and essential step of QC screening. Outliers can be caused by several reasons, including data entry errors, laboratory processing errors, or anomalous environmental events. Identifying and potentially removing or correcting outliers is a critical part of quality control. The functions in \CRANpkg{MassWateR} simply identify potential outliers to provide an opportunity for users to address these values. The decision on how to address outliers remains with the user and no automated tools are provided in the package to correct or remove outliers.

@@

\begin{center}\includegraphics[width=0.95\linewidth]{paper_files/figure-latex/unnamed-chunk-5-1} \end{center}

@@

The following describes the types of QC measurements that are evaluated by \CRANpkg{MassWateR}:

\begin{itemize}
\tightlist
\item
  \emph{Field Blanks}: Measurements in the field to assess lack of presence of a parameter, i.e., no measurement when the parameter is not present
\item
  \emph{Lab Blanks}: Laboratory samples to assess lack of presence of a parameter, i.e., no measurement when the parameter is not present
\item
  \emph{Field Duplicates}: Duplicate measurements in the field to assess similar values, i.e., precision between measurements is comparable
\item
  \emph{Lab Duplicates}: Duplicate laboratory samples to assess similar values, i.e., precision between samples is comparable
\item
  \emph{Calibration Checks}: Measurements in the field to verify an instrument returns a known quantity of a parameter, i.e., to verify accuracy of field sampling equipment
\item
  \emph{Lab Spikes}: Laboratory samples to verify if the processing equipment returns a known quantity of a parameter, i.e., to verify accuracy of laboratory equipment
\end{itemize}

Additionally, \CRANpkg{MassWateR} allows a user to include ``qualified samples'' in the results file. These samples are similar as those from the above criteria and are identified by the user as having special circumstances. For example, field duplicates are typically in pairs and the second measurement can be entered as a qualified sample in the same row of the first sample under the \texttt{"QC\ Reference\ Value"} column. This eliminates row repetition and facilitates preparation of the results file for the user.

@@

The QC report is built using several functions that can be used individually as needed. In particular, the \texttt{tabMWRacc()}, \texttt{tabMWRfre()}, and \texttt{tabMWRcom()} create \CRANpkg{flextable} (Gohel and Skintzos 2023) objects that can be viewed in RStudio and are compatible with Microsoft Word output. These functions are useful for understanding how the QC checks are created for the separate components of the QC report. For example, the \texttt{tabMWRacc()} function evaluates accuracy checks for field duplicates, lab duplicates, field blanks, lab blanks, and lab spikes/instrument checks for QC records in the results file based on DQOs in the accuracy file. The function can return a summary of all checks as follows (only ammonia and total phosphorus are shown for brevity):

@@

\begin{longtable}[c]{|p{1.47in}|p{0.55in}|p{1.08in}|p{0.88in}|p{0.72in}}

@@

\multicolumn{1}{>{\raggedright}m{\dimexpr 1.47in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{Type}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedright}m{\dimexpr 0.55in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{Parameter}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedright}m{\dimexpr 1.08in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{Number\ of\ QC\ Checks}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedright}m{\dimexpr 0.88in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{Number\ of\ Misses}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedright}m{\dimexpr 0.72in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{\%\ Acceptance}}} \\

@@

\multicolumn{1}{>{\raggedright}m{\dimexpr 1.47in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{Field\ Duplicates}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedright}m{\dimexpr 0.55in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedright}m{\dimexpr 1.08in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedright}m{\dimexpr 0.88in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedright}m{\dimexpr 0.72in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{}}} \\

@@

\multicolumn{1}{>{\raggedright}m{\dimexpr 1.47in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedright}m{\dimexpr 0.55in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{Ammonia}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedright}m{\dimexpr 1.08in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{4}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedright}m{\dimexpr 0.88in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{1}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedright}m{\dimexpr 0.72in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{75\ \%}}} \\

@@

\multicolumn{1}{>{\raggedright}m{\dimexpr 1.47in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedright}m{\dimexpr 0.55in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{TP}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedright}m{\dimexpr 1.08in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{5}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedright}m{\dimexpr 0.88in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{1}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedright}m{\dimexpr 0.72in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{80\ \%}}} \\

@@

\multicolumn{1}{>{\raggedright}m{\dimexpr 1.47in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{Lab\ Duplicates}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedright}m{\dimexpr 0.55in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedright}m{\dimexpr 1.08in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedright}m{\dimexpr 0.88in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedright}m{\dimexpr 0.72in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{}}} \\

@@

\multicolumn{1}{>{\raggedright}m{\dimexpr 1.47in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedright}m{\dimexpr 0.55in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{Ammonia}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedright}m{\dimexpr 1.08in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{10}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedright}m{\dimexpr 0.88in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{0}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedright}m{\dimexpr 0.72in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{100\ \%}}} \\

@@

\multicolumn{1}{>{\raggedright}m{\dimexpr 1.47in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedright}m{\dimexpr 0.55in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{TP}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedright}m{\dimexpr 1.08in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{16}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedright}m{\dimexpr 0.88in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{0}}} & \multicolumn{1}{!{\color[HTML]{666666}\vrule width 1pt}>{\raggedright}m{\dimexpr 0.72in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{7}{7}\selectfont{100\ \%}}} \\

@@

\begin{longtable}[c]{|p{0.55in}|p{0.60in}|p{0.27in}|p{0.50in}|p{0.53in}|p{0.45in}}

@@

@@

\begin{center}\includegraphics[width=0.95\linewidth]{paper_files/figure-latex/unnamed-chunk-17-1} \end{center}
